---
title: "Cyclistic Data Analysis About The Differences of Casual Riders And Members"
author: "Pedro Henrique Warken Ramos"
date: "2024-01-10"
output: pdf_document
---
# 1.Introduction
This data analysis report uses the historical data from all months of 2023 to determine
differences between how casual riders and annual members use Cyclistic bikes differently. With
this analysis it will be possible to create a market strategy to target casual riders and
encourage them to buy the annual membership. This report contains all the code used to analyse
the data provided along with data visualizations and explanations about the findings all the way
from creating the data frame, cleaning the data and manipulating it. At the end of the report there
will be the conclusions that can be made about the findings and a data driven market strategy to
convert casual riders into annual members.

# 2.Data analysis walkthrough

## 2.1 Create a data frame with all cyclistic data from 2023
```{r}
library(tidyverse)
data_list = list()
for (i in 1:12) {
  if (i < 10) {
    file_name = paste0("20230", i, "-divvy-tripdata.csv")
  } else {
    file_name = paste0("2023", i, "-divvy-tripdata.csv")
  }
  data_list[[i]] = read.csv(file_name)
}
cyclistic_2023 = do.call(rbind, data_list)
rm(data_list)
```
## 2.2 Verifying the integrity of the data:
```{r}
empty_counts = sapply(cyclistic_2023, function(x) sum(is.null(x) | x == ""))
```

With that is possible to conclude that there are lots of rows with missing values
for the station names and ids (for the end and start), also end_lat and end_lng
returned NA which should be further investigated.

```{r}
empty_counts = sapply(cyclistic_2023, function(x) sum(is.na(x)))
rm(empty_counts)
```

We can verify that there are several rows for which there is no end_lat and end_lng
which are likely bikes that were broken or stolen before reaching their destinations:

```{r}
na_end_positions = cyclistic_2023 %>%
  filter(is.na(end_lat))
```

Graph comparing the amount of problematic bikes from casual riders x annual members against
the total amount of each group:

```{r}
ggplot(na_end_positions, aes(
  x = member_casual,
  fill=member_casual,
)) + geom_bar() + labs(
  x = "User type",
  y = "Number of bike problems",
  title="Anual members with bike problems x\nCasual members with bike problems")
rm(na_end_positions)
ggplot(cyclistic_2023, aes(
  x = member_casual,
  fill=member_casual
)) + geom_bar() + labs(
  x = "User type",
  y = "Number of users",
  title="Total anual members x\nTotal casual members"
  )
```

With these graphs is clear that more bike problems occur in casual riders than
members, since there are more members in total but more casual riders with bike
problems. Nevertheless, with the data provided it is not clear exactly what happens
to the bikes or why, they might have broken, stolen or maybe the casual riders simply
did not understood correctly how to retrieve the bikes.

# 2.3 Manipulating the data
To further discover differences between casual riders and members we can analyze
calculated columns like total_trip_seconds and trip_distance_km, which represent
the total amount of seconds in each trip and the total distance in kilometers between
the starting station and the ending station respectively.
```{r}
# Total trip seconds:
library(lubridate)
library(dplyr)
manipulated_data = cyclistic_2023
rm(cyclistic_2023)
manipulated_data$started_at = ymd_hms(manipulated_data$started_at)
manipulated_data$ended_at = ymd_hms(manipulated_data$ended_at)
manipulated_data$total_trip_seconds = as.numeric(manipulated_data$ended_at - 
                                                   manipulated_data$started_at)
# Convert the numeric column to the 'mm:ss' format for better visualization
manipulated_data = manipulated_data %>%
  mutate(formatted_time = seconds_to_period(total_trip_seconds) %>%
           as.character() %>% 
           sprintf("%M:%S"))

# Add weekday
manipulated_data = manipulated_data %>%
  mutate(weekday_start = weekdays(started_at), weekday_end = weekdays(ended_at))

# Distance from bike retrieval to delivery:
# Function to calculate Haversine distance
haversine_distance = function(delta_latitude, delta_longitude) {
  # Radius of the Earth in kilometers
  earth_radius = 6371
  
  # Convert latitude and longitude differences to radians
  delta_lat_rad = delta_latitude * pi / 180
  delta_lon_rad = delta_longitude * pi / 180
  
  # Haversine formula
  a = sin(delta_lat_rad/2)^2 + cos((0) * pi / 180) * 
    cos((0 + delta_latitude) * pi / 180) * sin(delta_lon_rad/2)^2
  c = 2 * atan2(sqrt(a), sqrt(1 - a))
  distance = earth_radius * c
  
  return(distance)
}

distance_time_trips = manipulated_data %>%
  # Removing unfinished trips since they were already analyzed previously
  filter(!is.na(end_lat) & !is.na(end_lng)) %>%
  mutate(delta_lat = abs(end_lat - start_lat), 
         delta_lng = abs(end_lng - start_lng)) %>%
  mutate(trip_distance_km = haversine_distance(delta_lat, delta_lng))

rm(manipulated_data)
```

Preview of the first few rows of the manipulated data (only the columns added are
being shown for simplicity)
```{r}
head(distance_time_trips[, c(
  "total_trip_seconds", "formatted_time", "weekday_start", "weekday_end",
  "delta_lat", "delta_lng", "trip_distance_km"
  )])
```

# 2.4 Use additional calculated data to compare casual riders and members:
```{r}
grouped_data = distance_time_trips %>%
  group_by(member_casual) %>%
  summarize(average_time = mean(total_trip_seconds))

ggplot(grouped_data, aes(
  x = member_casual,
  y = average_time,
  fill = member_casual
)) + geom_bar(stat = "identity") +
  labs(
    x = "User type",
    y = "Average time in seconds",
    title="Total trip time comparison"
    )
```

```{r}
grouped_data = distance_time_trips %>%
  group_by(member_casual) %>%
  summarize(average_distance = mean(trip_distance_km))

ggplot(grouped_data, aes(
  x = member_casual,
  y = average_distance,
  fill = member_casual
)) + geom_bar(stat = "identity") + 
  labs(
    x = "User type",
    y = "Average distance in km",
    title = "Total distance trip comparison"
    )
rm(grouped_data)
```

```{r}
trips_0_distance = distance_time_trips %>%
  filter(trip_distance_km == 0)

ggplot(trips_0_distance, aes(
  x = member_casual,
  fill = member_casual
)) + geom_bar() + labs(
  x = "User type",
  y = "Count of trips endend in the same station",
  title = "Count of riders that started and\nended in the same station"
  )
rm(trips_0_distance)
```
Comparison of trips that started on weekdays between groups:
```{r}
trips_in_weekdays = distance_time_trips %>%
  group_by(member_casual) %>%
  summarize(
    total_trips = n(),  # Count total trips
    weekday_trips = sum(!(weekday_start %in% c("Saturday", "Sunday"))),
    weekend_trips = sum(weekday_start %in% c("Saturday", "Sunday"))
  )
rm(distance_time_trips)
print(trips_in_weekdays)

casual_data = subset(trips_in_weekdays, member_casual == "casual")
member_data = subset(trips_in_weekdays, member_casual == "member")

print(trips_in_weekdays)
casual_data = c(casual_data$weekday_trips, casual_data$weekend_trips)
member_data = c(member_data$weekday_trips, member_data$weekday_trips)
labels =
  c("Weekdays", "Weekends")
colors = c("orange", "blue")

par(mfrow = c(1, 2))
pie(
  casual_data, 
  labels = labels, 
  col = colors, 
  main = "Casual riders weekdays x\nweekends distribution")
pie(member_data, 
    labels = labels, 
    col = colors, 
    main = "Members weekdays x\nweekends distribution")
```

# 3. Conclusions and market strategy propose

Analyzing the first two graphs it is clear that more casual riders have problems with
the bikes than annual members although there are more annual members in total. Although
the problem that happens with the bikes is unknown with the data provided it is likely
a cause for the casual riders not wanting to continue using the app and buying an annual
subscription. This could be remedied by inserting clear instructions in the app and in each bike station
of how to get the bike, retrieve it and use it.

In the comparison of average time spent for each group the casual riders used the bikes considerably
longer the annual members. This is likely because many casual riders buy the single-ride pass, thus
they spend more time in the bikes to better harness the amount payed, since the charge is by ride and not
by time spent, whereas annual riders are can get as many bikes as they want because they already payed for
it. For the comparison of average distance between groups there wasn't a significant difference enough
to draw any conclusions about it.

About the analysis of rides that started and ended in the same station there is a small difference 
between each group, which likely just indicates that the users are testing the bike out for the first
time or just riding for fun which is more common in casual riders.

The last graph is the most significant finding. Opposite to what might be expected, there are more
casual riders using the bikes in weekdays than annual members. Since it is more common for people
to use bikes for fun in weekends it indicates that there are several casual riders that use the bikes
for locomotion for which an annual subscription would be more appropriate. This indicates two possible
options: the users fear committing for such a long time or the annual subscription does not present a
good enough discount compared to casual rides. This could be solved with a cancel and refund policy in
the first three months of use for example, solving the first issue and also implementing cheaper annual
plans that can be used only in weekdays, for example, or for a limited amount of days per week, solving
the second issue.

Using the data-driven marketing campaigns presented in this report the Cyclistic company will successfully
be able to convert a vast amount of casual riders into annual members providing becoming more lucrative 
and offering a better service to their clients.


